02_DESeq2

### 2020-02-18 analysis/02_DESeq2.Rmd
working on PRL results

```{r setup}
library(tidyverse)
library(DESeq2)
library(BiocParallel)
register(MulticoreParam(6))
library(caret)

@@

```{r PRL}
PRLdata <- read_csv("../results/PRLvsd.csv")
PRLsamples <- colData %>%
  mutate(samples = sapply(strsplit(as.character(V1), '.NYNO'), "[", 1))  %>%
  select(samples) 
PRLsamples

@@

dds <- DESeqDataSetFromMatrix(countData = newcountData,
                                colData = colDataPRL,
                                design = ~ tissue * sex * hiloPRL )
dds <- dds[rowSums(counts(dds) > 1) >= 10]  # filter more than sample with less 0 counts
print(dds)
print(dim(dds))
dds <- DESeq(dds, parallel = TRUE) # Differential expression analysis

vsd <- as.data.frame(assay(vst(dds, blind=FALSE)))
```

>>>>

```{r setup}
library(tidyverse)
library(DESeq2)
library(BiocParallel)
register(MulticoreParam(6))
library(caret) 
library(cowplot)

@@

```{r PRL}
PRLdata <- read_csv("../results/PRLvsd.csv")

PRLsamples <- colData %>%
  mutate(samples = V1)  %>%
  select(samples) 
PRLsamples

@@

dds <- DESeqDataSetFromMatrix(countData = newcountData,
                                colData = colDataPRL,
                                design = ~sex * hiloPRL )
dds <- dds[rowSums(counts(dds) > 1) >= 10]  # filter more than sample with less 0 counts
print(dds)
print(dim(dds))
dds <- DESeq(dds, parallel = TRUE) # Differential expression analysis

plotDispEsts( dds, ylim = c(1e-6, 1e1) )

ressex <- results( dds, contrast = c("sex", "male", "female") )
reshiloPRL <- results( dds, contrast = c("hiloPRL", "hi", "lo") )
hist( ressex$pvalue, breaks=20, col="grey" )
hist( reshiloPRL$pvalue, breaks=20, col="grey" )

restoDEGs <- function(res, up, down){
  DEGs <- data.frame(gene = row.names(res),
                        padj = res$padj, 
                        logpadj = -log10(res$padj),
                        lfc = res$log2FoldChange)
  DEGs <- na.omit(DEGs)
  DEGs <- DEGs %>%
    dplyr::mutate(direction = ifelse(DEGs$lfc > 0 & DEGs$padj < 0.1, 
                                     yes = up, no = ifelse(DEGs$lfc < 0 & DEGs$padj < 0.1, 
                                                           yes = down, no = "NS"))) %>% 
    dplyr::arrange(gene)
  DEGs$direction <- factor(DEGs$direction, levels = c(down, "NS", up)) 
  # write DEGsframe of only significant genes
  DEGs <- DEGs %>% dplyr::filter(direction != "NS")
  print(str(DEGs))
  #partialfilename = paste("_", down, "_", up, sep = "")
  #myfilename = paste0("../results/DESeq2/", mytissue, partialfilename, "_DEGs.csv")
  #write.csv(DEGs, myfilename, row.names = F)
  print(head(DEGs))
  return(DEGs)
}

sexdegs <- restoDEGs(ressex, "male", "female")
PRLdegs <- restoDEGs(reshiloPRL, "hi", "lo")

plotbarvolcano <- function(df, mylabels, mycolorname){
  p1 <- ggplot(df, aes(x = lfc, y = logpadj, color = direction)) +
    geom_point() +
    scale_color_manual(values = allcolors, name = mycolorname) +
    theme_B3() +
    theme(legend.position = "top") +
    labs(x = "Log fold change", y = "Log adj. p-value") +
    xlim(-5,5)

  p2 <- ggplot(df, aes(x = direction,  fill = direction)) +
    geom_bar() +
    theme_B3() +
    scale_fill_manual(values = allcolors) +
    theme_B3() +
    theme(legend.position = "none") +
    labs(x = mycolorname, y = "Total DEGs")

  p <- plot_grid(p1,p2, labels  = mylabels, label_size = 12)
  plot(p)
}

top <- plotbarvolcano(sexdegs, c("a","b"), "sex")
bottom <- plotbarvolcano(PRLdegs, c("c","d"), "prolactin expression")

plot_grid(top,bottom, nrow = 2)

head(PRLdegs)
head(sexdegs)

sexPRLdesg$direction.x <- as.character(sexPRLdesg$direction.x)

sexPRLdesg <- full_join(sexdegs, PRLdegs, by = "gene") %>%
  dplyr::select(gene, direction.x, direction.y) %>%
  tidyr::replace_na(list(direction.x = "no sex effect", direction.y = "no PRL effect"))
sexPRLdesg

ggplot(sexPRLdesg, aes(x = direction.x, fill = direction.y)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = allcolors, na.value= "#bdbdbd",
                    labels= c("low PRL", "hi PRL", "NS"),
                    name = "PRL DEGs") +
  scale_x_discrete(breaks = c("female", "male", NA),
                   labels = c("female", "male", "NS"),
                   name = "Sex DEGs")

interaction <- sexPRLdesg %>% filter(direction.x %in% c("male", "female") & 
                                       direction.y %in% c("hi", "lo"))
interaction

sexonly <- sexPRLdesg %>% filter(direction.x %in% c("male", "female") & 
                                       !direction.y %in% c("hi", "lo") )
sexonly
PRLonly <- sexPRLdesg %>% filter(!direction.x %in% c("male", "female")& 
                                       direction.y %in% c("hi", "lo"))
PRLonly

write.csv(sexPRLdesg, "../results/DEG-PRLsex.csv")
```
